---
title: "Draft Report"
author: "Jay Bendre"
date: "03/03/2022"
output:
  html_document:
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction

Covid-19 has managed to put this world to a complete stop and to date continues to be a thorn in our daily lives. With every day new variant and every day different mandates released by the local or federal governments has a direct correlation to the current low rates of cases and deaths being caused by the various variants. According to Stephanie Soucheray from CIDRAP, the covid cases have dropped by 40\% ([CIDRAP](https://www.cidrap.umn.edu/news-perspective/2022/02/covid-19-cases-drop-40-us)) in the last week alone from the first week of February to the second week of February. This has led to the loosening up of the mask mandates and other mandates by various states across the US. 

It is interesting to check for the effect of these mandates on the overall case fatality rate (CFR) whether they truly have been effective in slowing down the extent of the spread of the virus. Various countries employed policies on varying timeline, for this study we look at the top 10 most populous countries ie. United States of America, India, China, Indonesia, Pakistan, Nigeria, Brazil, Bangladesh, Russia and Mexico. We look at how closing of workplaces, banning international travel and financial compensations provided by the governments like debt reliefs had effect on reducing the fatality and understanding whether these mandates incentivized people from not stepping out of the house for work and making sure that the lockdown norms set up are not broken. 

The analysis occurs only pre and post implication of the mandates, we do not look at the various phases of mandates that were introduced, we just look at the effect of  introduction of the mandates and its effect in the reduction of the CFR.


## 2. Background 

## 3. Datasets 

The data being proposed to be used in the following project mainly comes from 2 sources:

#### * WHO Covid Data ([WHO Data](ttps://covid19.who.int/WHO-COVID-19-global-data.csv))

- The data obtained is gathered from the WHO Health Emergency Dashboard. It is an open-source dataset consisting of data pertaining to the number of cases and deaths that have occured over the past 2 years and arranged in a time-series capacity. 

- The table below shows all the important variables important in the dataset.

```{r echo = FALSE, message = FALSE, paged.print = TRUE, results = 'asis'}
var_list <- c("Date Reported", "Country Code", "Country", "WHO Region", "New Cases", "Cumulative Cases", "New Deaths", "Cumulative Deaths")
var_type <- c("Timestamp", "Factor", "Factor", "Factor", "Quantitative","Quantitative","Quantitative","Quantitative")
var_desc <- c("Timestamp of data collection", "Abbreviation of the Country name", "Country name", "Identifies various WHO regions", "Number of new cases", "Sum of cases till the given date", "Number of new deaths", "Sum of deaths till the given date")


var_table <- as.data.frame(list(var_list,var_type, var_desc))
colnames(var_table) <- c("Variable Name", "Variable Data type", "Variable Description")
kable(var_table) %>% kable_styling(position = "center")
```

#### * OxCGRT Dataset ([Oxford Data](https://github.com/OxCGRT/covid-policy-tracker))

- The data contained takes a look at many of the policies and mandates imposed by federal governments across entire nations. 

- Few of such mandates that could be interesting to look at are as follows 
```{r echo = FALSE, message = FALSE, paged.print = TRUE, results = 'asis'}

var_list <- c("Date","C1","C2","C6","C8","E1","E2","H3","H7", "V2A")
var_type <- c("Timestamp", "Factor", "Factor", "Factor","Factor","Factor","Factor","Factor","Factor","Factor")

var_desc <- c("Timestamp of data collection", "Indicator for School Closures", "Indicator for Workplace Closures","Indicator for Stay at Home order","Indicator for Internation Travel Restrictions","Indicator for Income Support","Indicator for Debt/Contract Relief","Indicator for Contact Tracing","Indicator for Vaccination Policy", "Indicator for Vaccination Availability")


var_table <- as.data.frame(list(var_list,var_type, var_desc))
colnames(var_table) <- c("Variable Name", "Variable Data type", "Variable Description")
kable(var_table) %>% kable_styling(position = "center")
```

## 4. Exploratory Data Analysis


## 4. Proposed Methodology 

#### a. Inferences from the Model 

#### b. Model Diagnostics 

## 5. Inferences 

## 6. Conclusion